Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-01 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
edited abstract and added one ref to intro

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions. However, they entail a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis function approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance and practical implementation of the method in relation to key factors such as the number of basis functions, domain of the prediction space, and smoothness of the latent function. We provide intuitive visualizations and recommendations for choosing the values of these factors, which make it easier for users to improve approximation accuracy and computational performance. We also propose diagnostics for checking that the number of basis functions and the domain of the prediction space are adequate given the data. The proposed approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\keywords{Gaussian process \and Low-rank Gaussian process \and Hilbert space methods \and Sparse Gaussian process \and Bayesian statistics \and Stan}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

Sparse GPs are based on low-rank approximations of the covariance matrix. The low-rank approximation with $m \ll n$ {\it inducing points} implies reduced memory requirements of $O(nm)$ and corresponding computational complexity of $O(nm^2)$. A unifying view on sparse GPs based on approximate generative methods is provided by \citet{quinonero2005unifying}, while a general review is provided by \citet{rasmussen2006gaussian}. \citet{Burt+Rasmussen+vanderWilk:2019} show that for regression with normally distributed covariates in $D$ dimensions and using the squared exponential covariance function, $M=O(\log^DN)$ is sufficient for an accurate approximation.
An alternative class of low-rank approximations is based on forming a basis function approximation with $m \ll n$ basis functions. The basis functions are usually presented explicitly, but can also be used to form a low-rank covariance matrix approximation. Common basis function approximations rest on the spectral analysis and series expansions of GPs \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}. Sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving}. Recently, \citet{hensman2017variational} presented a variational Fourier feature approximation for GPs that was derived for the Mat{\'e}rn class of kernels. Another related method for approximating kernels relies on random Fourier features  \citep{rahimi2008random,rahimi2009weighted}. Certain spline smoothing basis functions are equivalent to GPs with certain covariance functions \citep{wahba1990spline,Furrer+Nychka:2007}. Recent related work based on a spectral representation of GPs as an infinite series expansion with the Karhunen-Loève representation \citep[see, e.g.,][]{grenander1981abstract} is presented by \citet{JSSv090i10}.

{\color{blue} Aki: we could cite in the intro https://arxiv.org/abs/2111.01084 as a paper which
reviews alternative methods}

>>>>

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions. However, the direct implementation entails a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis function approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance, and practical recommendations for how to select the number of basis functions and the boundary factor. Intuitive visualizations and recommendations, make it easier for users to improve approximation accuracy and computational performance.  We also propose diagnostics for checking that the number of basis functions and the boundary factor are adequate given the data. The approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\keywords{Gaussian process \and Low-rank Gaussian process \and Hilbert space methods \and Sparse Gaussian process \and Bayesian statistics \and Stan}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

Sparse GPs are based on low-rank approximations of the covariance matrix. The low-rank approximation with $m \ll n$ {\it inducing points} implies reduced memory requirements of $O(nm)$ and corresponding computational complexity of $O(nm^2)$. A unifying view on sparse GPs based on approximate generative methods is provided by \citet{quinonero2005unifying}, while a general review is provided by \citet{rasmussen2006gaussian}. \citet{Burt+Rasmussen+vanderWilk:2019} show that for regression with normally distributed covariates in $D$ dimensions and using the squared exponential covariance function, $M=O(\log^DN)$ is sufficient for an accurate approximation.
An alternative class of low-rank approximations is based on forming a basis function approximation with $m \ll n$ basis functions. The basis functions are usually presented explicitly, but can also be used to form a low-rank covariance matrix approximation. Common basis function approximations rest on the spectral analysis and series expansions of GPs \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}. Sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving}. Recently, \citet{hensman2017variational} presented a variational Fourier feature approximation for GPs that was derived for the Mat{\'e}rn class of kernels. Another related method for approximating kernels relies on random Fourier features  \citep{rahimi2008random,rahimi2009weighted}. Certain spline smoothing basis functions are equivalent to GPs with certain covariance functions \citep{wahba1990spline,Furrer+Nychka:2007}. Recent related work based on a spectral representation of GPs as an infinite series expansion with the Karhunen-Loève representation \citep[see, e.g.,][]{grenander1981abstract} is presented by \citet{JSSv090i10}.
Yet another approach is to present Gaussian process using precision matrix, which is the inverse of the covariance matrix. If the precision matrix is sparse, computation taking the benefit of that sparsity can scale much better than $O(n^3)$. See, for example, review by \citet{Lindgren+Havard:2022:spde}.
